Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-18 ms/main.tex
edit result 2; merge appendix A back to results

\begin{corollary}[Symmetric horizontal transmission] 
\label{corollary:equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition.

@@

\autoref{fig:result2}a illustrates these conditions, which are obtained by setting $\alpha=0$ in Eqs.\ \ref{eq:vert_hori_equilibrium} and \ref{eq:cost_boundaries}. If there is horizontal transmission bias for cooperation ($T_A>T_B$) then cooperation can evolve from any initial frequency if the cost is low enough, $c<\gamma_1$ (green area below solid line). If the cost is not low enough, cooperation can also evolve if its initial frequency is high enough and the cost is not too high, $c<\gamma_2$ (blue area between solid and dashed lines).

@@

\begin{corollary}[Perfect social association of transmission and cooperation]
\label{corollary:alpha1}
With perfect social association ($\alpha=1$), the only equilibria are the fixation states, $\tilde{p}=0$ and $\tilde{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\tilde{p}'>\tilde{p}$) if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
This condition is obtained from inequality \ref{eq:vert_hori_alpha1_condition_proof}, and the right-hand side equals $\gamma_1$ when $\alpha=1$ (\autoref{eq:cost_boundaries}).
Perfect social association (horizontal transmission always occurs during the cooperative interaction) is also assumed in the model studied by \citet{lewin2017microbes}, and therefore this corollary is equivalent to the result in their eq.~1.

@@

\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough; $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

@@

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
To proceed,  note that 
\autoref{eq:horizontal} gives $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
\autoref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
\autoref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
%TODO YR cleanup  
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}),
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

%We start with the simple case of symmetrical horizontal transmission, $T=T_A=T_B$ and apply \autoref{eq:horizontal}, \autoref{eq:nextgen_adults_slimpify}, and \autoref{eq:nextgen_parents_simplified} to obtain 
%\begin{equation} \label{eq:equal_horizontal_transmission}
%\begin{aligned}
%  f(\hat{p}) &= 
%  \bar{w}(\hat{p}' - \hat{p}) = \\
%  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].\end{aligned}
%\end{equation}
%A detailed explain of how we get eq.~\ref{eq:equal_horizontal_transmission} can be found in appendix B.
%The equilibria are solutions of $f(\hat{p})=0$, or $\hat{p}' = \hat{p}$.
%It is easy to verify that fixation of either phenotype, $\hat{p} =  0$ and $\hat{p} = 1$, is an equilibrium.
%Since the derivative of $f'(\hat{p}$ is 
%\begin{equation}
%f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
%\end{equation}
%then the condition for local stability of $\hat{p}=1$ is
%\begin{equation} \label{eq:derivative_of_phattag-phat}
%	f'(1) =	-\alpha bvT + cv(1-T) < 0,
%\end{equation}
%which gives the following result.
%
%\textbf{Result 3: Oblique and vertical transmission with symmetric horizontal transmission.}
%If horizontal transmission is symmetric, $T = T_A = T_B$, and if
%\begin{equation} \label{eq:oblique_and_vertic_result2}
%  %\frac{b}{c}>\frac{1-T}{\alpha T}
%	c < b \cdot \frac{\alpha T}{1-T},
%\end{equation}
%then fixation of the cooperator phenotype $A$ is locally stable.
%The same condition was given in Corollary 1.1, \autoref{eq:equal_transmission}.

%We now turn to the general case where $T_A \neq T_B$. 
Applying Eqs.\ \ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} gives the function $f(\hat{p})$ (see Appendix~\autoref{sec:appendixA}):

\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) , \\
\beta_2 &= -\beta_1 -\beta_3 ,  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixB}), where
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{corollary:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixA}
We want to find the frequency of juveniles with phenotype $A$ in next generation $\hat{p}'$ as a function of frequency
of juveniles with phenotype $A$ in the current generation $\hat{p}$.
Starting from \autoref{eq:horizontal},
\begin{equation}\label{eq:appendix_b_1}
  \hat{p}' = v \tilde{p}' + (1-v) p',
\end{equation}
 substitute for $p'$ using \autoref{eq:nextgen_adults_slimpify} and for $\tilde{p}'$ using \autoref{eq:nextgen_parents_simplified}, we have
\begin{equation}\label{eq:appendix_b_2}
  \begin{aligned}
  \hat{p}'  = & \frac{v}{\bar{w}}\left\{\hat{p}^2(1+b-c)\left[1-(1-\hat{p})(1-\alpha)T_B)\right]\right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1-c)\left[ \hat{p}(1-\alpha)T_B + 1 - T_B \right] \right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1+b)\left[\hat{p}(1-\alpha) + \alpha \right]T_A \right\} \\
  & + \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \\
  & + (1-v)\hat{p}^2(T_B-T_A) + (1-v)\hat{p}(1+T_A-T_B),
\end{aligned}
\end{equation}
where $\bar{w} = 1 + \hat{p}(b-c)$. 
Define $g(\hat{p})$ as
\begin{equation} \label{eq:appendix_define_f}
\begin{aligned}
      g(\hat{p}) &= \bar{w}(\hat{p}' - \hat{p})
\end{aligned}
\end{equation}
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics,  \autoref{eq:appendix_define_f} simplifies to 
eqs.~\ref{eq:general_case_polynomial}-\ref{eq:polynomial_coefficients}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixB}

>>>>

\begin{corollary}[Symmetric horizontal transmission] 
\label{corollary:equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition. Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.

@@

\autoref{fig:result2}a illustrates these conditions, which are obtained by setting $\alpha=0$ in Eqs.\ \ref{eq:vert_hori_equilibrium} and \ref{eq:cost_boundaries}. If there is horizontal transmission bias for cooperation ($T_A>T_B$) then cooperation can evolve from any initial frequency if the cost is low enough, $c<\gamma_1$ (green area below solid line).
If the cost is not low enough, cooperation can still evolve if its initial frequency is high enough and the cost is not too high, $c<\gamma_2$ (blue area between solid and dashed lines).

@@

\begin{corollary}[Perfect social association of transmission and cooperation]
\label{corollary:alpha1}
With perfect social association ($\alpha=1$), the only equilibria are the fixation states, $\tilde{p}=0$ and $\tilde{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\tilde{p}'>\tilde{p}$) if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
This condition is obtained from inequality \ref{eq:vert_hori_alpha1_condition_proof}, and the right-hand side equals $\gamma_1$ when $\alpha=1$ (\autoref{eq:cost_boundaries}).
Perfect social association (horizontal transmission always occurs during the social interaction) is also assumed in the model studied by \citet{lewin2017microbes}, and therefore this corollary is equivalent to their eq.~1.

@@

\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough, $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

@@

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
In the following, we find an equation for the frequency of phenotype $A$ among juveniles in the next generation $\hat{p}'$ as a function of the frequency in the current generation,
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.

Starting from \autoref{eq:horizontal}, we substitute \autoref{eq:nextgen_adults_slimpify} for $p'$ and  \autoref{eq:nextgen_parents_simplified} for $\tilde{p}'$.
We therefore have
\begin{equation}
  \begin{aligned}
  \hat{p}'  = & \frac{v}{\bar{w}}\left\{\hat{p}^2(1+b-c)\left[1-(1-\hat{p})(1-\alpha)T_B)\right]\right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1-c)\left[ \hat{p}(1-\alpha)T_B + 1 - T_B \right] \right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1+b)\left[\hat{p}(1-\alpha) + \alpha \right]T_A \right\} \\
  & + \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \\
  & + (1-v)\hat{p}^2(T_B-T_A) + (1-v)\hat{p}(1+T_A-T_B),
\end{aligned}
\end{equation}
where $\bar{w} = 1 + \hat{p}(b-c)$. 
We define $f(\hat{p}) = \bar{w}(\hat{p}' - \hat{p})$.
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics,  this simplifies to
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) , \\
\beta_2 &= -\beta_1 -\beta_3 ,  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$. 
Thus, $f(\hat{p})$ becomes a quadratic polynomial,
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixA}), where
$f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big]$, so that
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{corollary:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixA}
